---
title: "STAT545 Assignment 4"
author: "Alex"
date: "October 9, 2018"
output: 
  github_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)

```

## Cheatsheet for Data Reshaping Using Tidyr

Two of the most useful functions in tidyr for data reshaping are the `gather()` and `spread()` functions. Let's take a quick look at how to use them!

### gather()

This function is helpful when you have multiple datasets that you want to combine into a single data frame with nice clean columns. 

#### Example

A good practical example is from our Lecture 09, which looks at cleaning the untidy dataset that we get from haphazardly combining three lotr data files:

```{r, message=FALSE}
lotr1 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv")
lotr2 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv")
lotr3 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv")
```

```{r}
untidy <- bind_rows(lotr1,lotr2,lotr3) # stack the datasets on top of each other
untidy %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
```

Here we see that we have 4 columns *Film, Race, Female*, and *Male*. This isn't toooooo untidy, but it makes more sense to aggregate the Male and Female columns into one variable since they are just two halves of one larger variable that measures *Words*, and just have another variable *Gender* to retain the gender information. This makes it a lot easier for other functions to run on this dataset.

We can use gather() to merge the Female and Male columns into one column called *Word*, with the Male and Female information coded in another column called *Gender*:

```{r}
tidy <- gather(untidy, key="Gender", value="Words", Female, Male)
tidy %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
  
```


#### Summary Table for gather()

gather(data, key, value, ...)

| Argument | Description |
|----------|-------------------------------------------------|
| data | The untidy dataset you want to tidy up |
| key | The name you want for the variable that will code the columns you want to aggregate together |
| value | The column name you want for the aggregated data |
| ... | Names of all the columns that you want to aggregate |

### spread()

spread() can be thought of as doing the opposite of what gather() does. If we have data in one column that we would rather have separated into multiple columns based on its value on some variable, then we use spread()!


#### Example

We can use the tidy lotr dataset from above as an example. Suppose we wanted to have a separate column of words spoken for each of the Races *Elf, Hobbit,* and *Man*. We can do this by:

```{r}
tidy %>%
  spread(key = Race, value = Words) %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
```

We see now that the *Words* column has been spread into three different columns, one for each of the 3 different Races! This is very useful when you wanted to quickly calculate summary statistics based on a subset of your data and you have too many different subsets to efficiently do it manually!
 
#### Summary

spread(data, key, value)

| Argument | Description |
|----------|-------------------------------------------------|
| data | The dataset you want to work on |
| key | The varaible whose values you wish to use to spread the values of another variable over |
| value | The variable whose values you want to spread over your *key* variable |



## Cheatsheet for the family of join() functions

### Data Prep

As may be apparent from my first assigment, I'm really interested in jazz! In order to look at how the join() functions work, let's first prep a dataset that consists of Name, Age, Instrument, Most Prominent Era:

```{r}
jazzgreats <- "
Name, Age, Instrument, Genre
Oscar Peterson, 82, Piano, Swing
Count Basie, 79, Piano, Big Band
Miles Davis, 65, Trumpet, Modal
John Coltrane, 40, Saxophone, Hard Bop
Duke Ellington, 75, Piano, Big Band
Buddy Rich, 69, Drums, Big Band
Charlie Parker, 34, Saxophone, Bebop
Bill Evans, 51, Piano, Modal
Chick Corea, 77, Piano, Fusion
Ella Fitzgerald, 79, Vocals, Swing
Ornette Coleman, 85, Saxophone, Free Jazz
"

jazzgreats <- read_csv(jazzgreats, skip = 1) 

jazzgreats %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
  
```

Great! Now suppose we have another dataset that contains different jazz genres along with the starting year for each genre:

```{r}
genre_years <- "
Genre, Year
Dixieland, 1920
Swing, 1930
Big Band, 1930
Bebop, 1945
Modal, 1950
Hard Bop, 1955
Fusion, 1970
"

genre_years <- read_csv(genre_years, skip = 1)

genre_years %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))

```

The information in the `genre_years` data frame would be very useful to have in our `jazzgreats` data frame. Let's look at how we can use different join() functions to accomplish this, and how each one differs!

### Different join functions
#### inner_join()

inner_join(x,y) returns all rows in x that have matching values in y, then appends all applicable rows from y to x.

```{r}
inner_join(jazzgreats, genre_years) %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
```

All entries in `jazzgreats` that have a Genre that also appears in `genre_years` have been returned. We see that with this join, we have lost Ornette Coleman from `jazzgreats` since *Free Jazz* was not a genre in `genre_years`.


#### semi_join()

semi_join(x,y) is like inner_join(x,y), except at the end it returns only the rows that belonged to x.

```{r}
semi_join(jazzgreats, genre_years) %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
```

We see that we get exactly the same result as with inner_join, except the *Years* column has not been appended to `jazzgreats`.

#### left_join()

left_join(x,y) returns all rows from x with all columns from both x and y. This differs from inner_join(x,y) in that here if there is no corresponding row for a certain value in x, we will get NAs instead of nothing at all.

```{r}
left_join(genre_years, jazzgreats) %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
```

We see that even though there were no artists in `jazzgreats` that had Dixieland as their most prominent genre, we still retain *Dixieland* as a variable since it was in `x = genre_years`

#### anti_join()

anti_join(x,y) keeps all rows from x that do not have matching values in y.

```{r}
anti_join(jazzgreats, genre_years) %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))

anti_join(genre_years, jazzgreats) %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
```

We see that we get different results depending on which order we put `jazzgreats` and `genre_years`. This is because each dataset has one row that does not have matching values in the other dataset. With `x = jazzgreats` we see that we are left with Ornette Coleman, since his genre of *Free Jazz* is not found in `genre_years`. With `x = genre_years` we are left with Dixieland, since *Dixieland* is the only genre not found in `jazzgreats`.

#### full_join()

full_join(x,y) returns all row and columns from both x and y. This basically aggregates all data from both data frames x and y. NAs are added where there is no matching data.

```{r}
full_join(jazzgreats, genre_years) %>%
  kable("html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", full_width = F))
```

We see that we do indeed get back every row and column from both data frames. In the cases of *Ornette Coleman* from `jazzgreats` and *Dixieland* from `genre_years`, we see that NAs have been filled in.

### Summary of join functions

Let's summarize the join functions in plain english:

| Function | Description |
|----------|-------------------------------------------------|
| inner_join(x,y) | Return all rows from x that have matching values in y, along with the respective data from every column in y |
| semi_join(x,y) | Same as inner_join(x,y) but only returns data from x at the end |
| left_join(x,y) | Similar to inner_join(x,y), but instead of omitting rows in y that do not have matching rows in x, we create a new row with NAs as needed |
| anti_join(x,y) | Returns all rows in x that do not have matching rows in y |
| full_join(x,y) | Similar to a left_join(x,y), but this keeps all data from all rows and columns in both x and y, then fills in NAs as needed |







